---
title: "BMI 714 Final Project!"
author: "Neha Dantuluri"
date: "December 16, 2024"
output:
  html_document: 
toc: true
toc_depth: 3
toc_float: true
pdf_document: default
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_engines$set(txt = function(options) {
  code <- paste(options$code, collapse = "\n")
  knitr::engine_output(options, code, NULL)
  })
```


## 1: Exploratory Analysis and QC of Data

1.1 Data Transformation + Assessing/removing/imputing missing values

```{r}
library(tidyverse)
library(haven)

demographics_data <- read_xpt("/Users/nehadantuluri/Downloads/P_DEMO.XPT")
depression_data <- read_xpt("/Users/nehadantuluri/Downloads/P_DPQ.XPT")
occupation_data <- read_xpt("/Users/nehadantuluri/Downloads/P_OCQ.XPT")
health_insurance_data <- read_xpt("/Users/nehadantuluri/Downloads/P_HIQ.XPT")
hospital_use_data <- read_xpt("/Users/nehadantuluri/Downloads/P_HUQ.XPT")
food_security_data <- read_xpt("/Users/nehadantuluri/Downloads/P_FSQ.XPT")

nhanes_data <- reduce(list(demographics_data, depression_data, occupation_data, health_insurance_data, hospital_use_data, food_security_data), 
                      merge, by = "SEQN", all = TRUE)

selected_nhanes_variables <- nhanes_data %>%
  select(SEQN, DPQ010, RIAGENDR, RIDAGEYR, RIDRETH3, DMDMARTZ, DMDEDUC2, INDFMPIR, OCD150, HIQ011, HUQ090, FSDAD) %>%
  rename(
    respondent_ID = SEQN,
    depression_score = DPQ010,
    gender = RIAGENDR,
    age = RIDAGEYR,
    race_ethnicity = RIDRETH3,
    marital_status = DMDMARTZ,
    education_level = DMDEDUC2,
    income2poverty_ratio = INDFMPIR,
    employment_measures = OCD150,
    health_insurance = HIQ011,
    mental_health_seen = HUQ090,
    food_security_status = FSDAD) %>% 
  filter(
    age >= 20 & age <= 49,                 
    gender == 2,                            
    depression_score %in% c(0,1,2,3),      
    race_ethnicity  %in% c(1,2,3,4,6,7),    
    marital_status %in% c(1,2,3),          
    education_level  %in% c(1,2,3,4,5),   
    income2poverty_ratio <= 4.98,          
    health_insurance %in% c(1,2),         
    mental_health_seen %in% c(1,2),        
    employment_measures %in% c(1,2,3,4),      
    food_security_status %in% c(1,2,3,4)) %>% 
  mutate(
     employment_status= case_when(
      employment_measures %in% c(1, 2) ~ "Employed", 
      employment_measures == 3 ~ "Looking for work",
      employment_measures == 4 ~ "Not working"))

write.csv(selected_nhanes_variables, "/Users/nehadantuluri/Downloads/nhanes_selected_data.csv", row.names = FALSE)
nhanes_clean <- read.csv("/Users/nehadantuluri/Downloads/nhanes_selected_data.csv")
head(nhanes_clean)
```

1.2 Plots showing relationship between variables of interest. 
```{r}
library(ggplot2)

# categorical variable x numeric variable plots
ggplot(nhanes_clean, aes(x = factor(depression_score), y = age)) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Age vs. Depression Score (Jittered)", x = "Depression Score", y = "Age") +
  theme_minimal()

ggplot(nhanes_clean, aes(x = factor(depression_score), y = income2poverty_ratio)) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Income-to-Poverty Ratio vs. Depression Score (Jittered)", x = "Income-to-Poverty Ratio", y = "Age") +
  theme_minimal()

# categorical variable x categorical variable plots
ggplot(data = nhanes_clean, aes(x = factor(depression_score), fill = factor(race_ethnicity))) +
  geom_bar(position = "fill") + 
  labs(title = "Proportion of Race/Ethnicity Across Depression Severity", x = "Depression Score", y = "Proportion", fill = "Race/Ethnicity") +
  theme_minimal()

ggplot(data = nhanes_clean, aes(x = factor(depression_score), fill = factor(marital_status))) +
  geom_bar(position = "fill") + 
  labs(title = "Proportion of Marital Status Across Depression Severity", x = "Depression Score", y = "Proportion", fill = "Marital Status") +
  theme_minimal()

ggplot(data = nhanes_clean, aes(x = factor(depression_score), fill = factor(education_level))) +
  geom_bar(position = "fill") + 
  labs(title = "Proportion of Education Level Across Depression Severity", x = "Depression Score", y = "Proportion", fill = "Education Level") +
  theme_minimal()

ggplot(data = nhanes_clean, aes(x = factor(depression_score), fill = factor(employment_status))) +
  geom_bar(position = "fill") + 
  labs(title = "Proportion of Employment Status Across Depression Severity", x = "Depression Score", y = "Proportion", fill = "Employment Status") +
  theme_minimal()

ggplot(data = nhanes_clean, aes(x = factor(depression_score), fill = factor(health_insurance))) +
  geom_bar(position = "fill") + 
  labs(title = "Proportion of Health Insurance Status Across Depression Severity", x = "Depression Score", y = "Proportion", fill = "Health Insurance Status") +
  theme_minimal()

ggplot(data = nhanes_clean, aes(x = factor(depression_score), fill = factor(mental_health_seen))) +
  geom_bar(position = "fill") + 
  labs(title = "Proportion of Seen by Mental Health Professional Across Depression Severity", x = "Depression Score", y = "Proportion", fill = "Seen by a Mental Health Professional") +
  theme_minimal()

ggplot(data = nhanes_clean, aes(x = factor(depression_score), fill = factor(food_security_status))) +
  geom_bar(position = "fill") + 
  labs(title = "Proportion of Food Security Status Across Depression Severity", x = "Depression Score", y = "Proportion", fill = "Food Security Status") +
  theme_minimal()


# used this resource for ggplot2: https://ggplot2.tidyverse.org/reference/theme.html 
# used this resource for geom_jitter: https://thomasleeper.com/Rcourse/Tutorials/jitter.html 
```

1.3 Check correlations/collinearity
```{r}
cor(nhanes_clean$depression_score, nhanes_clean$age, method = "spearman") # value is close to 0, no significant correlation
cor(nhanes_clean$depression_score, nhanes_clean$income2poverty_ratio, method = "spearman") # weak negative correlation, higher income may may slightly influence lower depression scores 
cor(nhanes_clean$depression_score, nhanes_clean$race_ethnicity, method = "spearman") # value is close to 0, no significant correlation
cor(nhanes_clean$depression_score, nhanes_clean$marital_status, method = "spearman") # value is close to 0, no significant correlation 
cor(nhanes_clean$depression_score, nhanes_clean$education_level, method = "spearman") # value is close to 0, no significant correlation 
cor(nhanes_clean$depression_score, nhanes_clean$employment_status, method = "spearman") # weak positive correlation, employment status may slightly influence depression scores
cor(nhanes_clean$depression_score, nhanes_clean$health_insurance, method = "spearman") # value is close to 0, no significant correlation 
cor(nhanes_clean$depression_score, nhanes_clean$mental_health_seen, method = "spearman") # weak negative correlation, more mental_health_seen may slightly influence lower depression scores
cor(nhanes_clean$depression_score, nhanes_clean$food_security_status, method = "spearman") # weak positive correlation, food security status may slightly influence depression scores
```

1.4 Scale predictors as needed
```{r}
nhanes_clean <- nhanes_clean %>%
  mutate(
    age_scaled = scale(age),
    income2poverty_ratio_scaled = scale(income2poverty_ratio))
```


1.5 Code categorical variables appropriately
```{r}
nhanes_clean <- nhanes_clean %>%
  mutate(
    depression_score = factor(depression_score, levels = c(0, 1, 2, 3), 
                              labels = c("Not Difficult", "Somewhat Difficult", "Very Difficult", "Extremely Difficult")),
    gender = factor(gender, levels = c(1, 2), labels = c("Male", "Female")),
    race_ethnicity = factor(race_ethnicity, levels = c(1, 2, 3, 4, 6, 7), 
                            labels = c("Mexican American", "Other Hispanic", "Non-Hispanic White", 
                                       "Non-Hispanic Black", "Non-Hispanic Asian", "Other/Multi-Racial")),
    marital_status = factor(marital_status, levels = c(1, 2, 3), 
                            labels = c("Married/Living with Partner", "Widowed/Divorced/Separated", "Never Married")),
    education_level = factor(education_level, levels = c(1, 2, 3, 4, 5), 
                             labels = c("Less than 9th Grade", "9-11th Grade", "High School Graduate", 
                                        "Some College/AA Degree", "College Graduate or Above")),
    health_insurance = factor(health_insurance, levels = c(1, 2), labels = c("Yes", "No")),
    mental_health_seen = factor(mental_health_seen, levels = c(1, 2), labels = c("Yes", "No")),
    food_security_status = factor(food_security_status, levels = c(1, 2, 3, 4), 
                                  labels = c("Full Food Security", "Marginal Food Security", 
                                             "Low Food Security", "Very Low Food Security"))
  )
```



## 2: CHOOSE A REGRESSION OR CLASSIFICATION MODEL

2.1 Choose Model, Evaluate Assumptions, & Justify Decisions


## 3: RUN THE MODEL

```{r}
library(MASS)
```

3.1 Run initial model

```{r}
# ordinal logistic regression
model <- polr(depression_score ~ income2poverty_ratio + employment_status + 
                     marital_status + health_insurance + mental_health_seen + 
                     education_level + race_ethnicity +
                     food_security_status, data = nhanes_clean, method = "logistic")
summary(model)
```

3.2 Variable Selection

```{r}
# Stepwise Variable Selection (AIC)
selected_model <- stepAIC(model, direction = "both", trace = TRUE)
summary(selected_model)
```

3.3 Run model (based on variable selection)

```{r}
model_2 <- polr(depression_score ~  income2poverty_ratio + health_insurance + 
                mental_health_seen + food_security_status, 
                data = nhanes_clean, method = "logistic")
summary(model_2)
```


## 4: EVALUATE MODEL FIT
4.1 Choose, calculate, and appropriately interpret a metric 
4.2 Justify the metric you chose based on the goal of your model

## 5: COMPARE YOUR MODEL TO ONE WITH DIFFERENT VARIABLES
5.1 Justify alternate model chosen

5.2 Run alternate model
```{r}
alternate_model <- polr(depression_score ~ income2poverty_ratio + marital_status +  
                          education_level + race_ethnicity + food_security_status,
                     data = nhanes_clean, method = "logistic")
```

5.3 Calculate a metric to compare models

```{r}
anova(model_2, alternate_model, test = "Chisq")
```


## 6: CONDUCT ONE FOLLOW UP ANALYSIS

6.1 Choose an analysis which fits with the over goal of your model

```{r}
followup_model <- polr(depression_score ~ income2poverty_ratio + food_security_status + marital_status + education_level, 
                        data = nhanes_clean[nhanes_clean$age >= 20 & nhanes_clean$age <= 34,], method = "logistic")
summary(followup_model)
```


6.2 Appropriately interpret the results



